NumPy---Array Programming in Python

### 2020-02-01 body.tex
One proliferation section change; some textual things and comments
We're getting to the stage where it's hard to find obvious issues;
a few things I wanted to change but I thought it'd be better to
add comments for discussion instead.

% KJM: I moved this here to get the placement right 
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
  \caption{\textbf{Fundamental Array Concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with various types of arguments.
    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    \textbf{d,} Vectorization efficiently applies operations to groups of elements
    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
    \textbf{f,} Reduction operations act along one or more axes. In this
    example, a three-dimensional array is shown to be summed along various single
    axes to produce two-dimensional results, or along two axes consecutively to
    produce a one-dimensional result.
    \textbf{g,} Example NumPy code, illustrating some of these concepts.
   }
  \label{fig:array-concepts}
\end{figure*}

@@

The NumPy array is a data structure that efficiently stores and accesses
multidimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, that
enables a wide variety of scientific computation.
It consists of a pointer to memory, along with metadata used to interpret the
data stored there, notably {\em data type}, {\em shape}, and {\em strides}
(Fig.~\ref{fig:array-concepts}a).

@@

\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights just a few of the essential tools and libraries in the vast
   scientific Python ecosystem.  Most of the ecosystem builds on NumPy arrays to
   provide a rich and powerful collection of array-aware functionality.
  }
  \label{fig:ecosystem}
\end{figure}

Many research groups have designed large,
complex scientific libraries, which add specialized functionality for
solving problems in specific domains.
For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration for radio interferometry imaging, analysis, and simulation
relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image
reconstruction.
SciPy is used to provide tools used in general image processing such as
filtering and image alignment, while scikit-image, an image processing
library that extends SciPy, provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling general optimization tasks.
NetworkX \cite{SciPyProceedings_11}, a package for complex
network analysis, is used for analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including for the generation of the final image of the black hole.

@@

NumPy provides in-memory, multidimensional, homogeneously typed (i.e.,
single pointer and strided) arrays on CPUs.  It runs on machines
ranging from embedded devices to the world's largest supercomputers,
with performance approaching that of compiled languages.  For the
largest part of its existence, NumPy covered the vast majority of
array computation use cases.

@@

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with arrays is experimental by nature
and, in fact, several promising libraries---such as Caffe, Theano, and
biggus---have already ceased development. And each time that a user
decides to try a new technology or backend library, they have to
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

@@

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.
Thanks to these developments users of Dask and NumPy arrays, for instance, are now capable
of scaling their computation to the GPU using Dask's CuPy integration with
only minimal code changes.

@@

NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

This culture of using best practices for producing reliable scientific software
has been adopted by the ecosystem of libraries that build on NumPy.
For example, in a recent award given by the Royal Astronomical Society to
Astropy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists
with experience in professional-standard software development practices
including use of version control, unit testing, code review and issue tracking
procedures. This is a vital skill set for modern researchers that is often
missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through formal courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.

>>>>

% KJM: I moved this here to get the placement right 
% RG: minor comment on the example code: would be better to not copy IPython
%     oddities but align output on brackets and In/Out on ":"
\begin{figure*}[h]
  \centering
  \includegraphics[width=\textwidth]{static/sketches/array-concepts}   
  \caption{\textbf{Fundamental Array Concepts.}
    \textbf{a,} The NumPy array data structure and its associated metadata fields.
    \textbf{b,} Indexing an array with various types of arguments.
    \textbf{c,} Indexing with arrays, which broadcast the indexing arguments before performing the lookup.
    \textbf{d,} Vectorization efficiently applies operations to groups of elements
    \textbf{e,} Broadcasting in scalar addition, and in the division of two-dimensional arrays.
    \textbf{f,} Reduction operations act along one or more axes. In this
    example, a three-dimensional array is shown to be summed along various single
    axes to produce two-dimensional results, or along two axes consecutively to
    produce a one-dimensional result.
    \textbf{g,} Example NumPy code, illustrating some of these concepts.
   }
  \label{fig:array-concepts}
\end{figure*}

@@

The NumPy array is a data structure that efficiently stores and accesses
multidimensional arrays \cite{vanderwalt2011numpy}, also known as tensors, and
enables a wide variety of scientific computation.
It consists of a pointer to memory, along with metadata used to interpret the
data stored there, notably {\em data type}, {\em shape}, and {\em strides}
(Fig.~\ref{fig:array-concepts}a).

@@

% RG: (to discuss) this figure plus associated text says that the layers about
%     Foundation build on top of that whole foundation. In reality that's not
%     true - everything builds on NumPy, almost everything on SciPy, and very
%     little on Matplotlib/IPython/Jupyter. Perhaps "interactive environment"
%     should get its own section instead.
%     Also, I think Pandas can be added next to statsmodels as statistics. That
%     avoids explaining dataframes; now it looks like a serious omission.
%     I'd also be in favor of adding Spyder in the text; scientific computing
%     beyond exploratory work is better done in an IDE.
\begin{figure}
  \centering
  \includegraphics[width=.45\textwidth]{static/ecosystem}
  \caption{\textbf{Scientific Python Ecosystem.}
   This figure highlights just a few of the essential tools and libraries in the vast
   scientific Python ecosystem.  Most of the ecosystem builds on NumPy arrays to
   provide a rich and powerful collection of array-aware functionality.
  }
  \label{fig:ecosystem}
\end{figure}

Many research groups have designed large,
complex scientific libraries, which add specialized functionality for
solving problems in specific domains.
For example, the \code{eht-imaging} library developed by the Event Horizon Telescope 
collaboration for radio interferometry imaging, analysis, and simulation
relies on many components of the scientific Python ecosystem.
NumPy arrays are used to store and manipulate numerical data at every step
in the processing chain: from raw data through calibration and image
reconstruction.
SciPy is used to provide tools used in general image processing such as
filtering and image alignment, while scikit-image, an image processing
library that extends SciPy, provides higher-level functionality such as
edge filters and Hough transforms.
Other components of SciPy are used throughout the library, such as the
\code{scipy.optimize} module for handling mathematical optimization tasks.
NetworkX \cite{SciPyProceedings_11}, a package for complex
network analysis, is used for analysis of consistency among image
comparisons.
The community-developed astronomy library, Astropy \cite{astropy:2013, astropy:2018},
is used throughout \code{eht-imaging}, primarily to handle standard
astronomical file formats and time/coordinate transformations.
Matplotlib is used for visualizing data throughout the analysis pipeline,
including for the generation of the final image of the black hole.

@@

NumPy provides in-memory, multidimensional, homogeneously typed
arrays on CPUs.  It runs on machines
ranging from embedded devices to the world's largest supercomputers,
with performance approaching that of compiled languages.  For the
largest part of its existence, NumPy covered the vast majority of
array computation use cases.

@@

Such libraries often mimic the NumPy API, because it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray.
But exploring new ways of working with arrays is experimental by nature
and, in fact, several promising libraries---such as Theano and Caffe---have
already ceased development. And each time that a user
decides to try a new technology, they have to
change import statements and ensure that the new library implements
all the parts of the NumPy API they currently use.

@@

To facilitate this \emph{interoperability}, NumPy provides
``protocols'' (or contracts of operation), that allow for these arrays to be
passed to NumPy functions (Fig.~\ref{fig:duck-arrays}).
NumPy, in turn, dispatches operations to the originating library, as required.

Thanks to these developments users are now capable, for example, of scaling
their computation to distributed systems with Dask. The protocols
\textit{compose} well, so it is even possible to take NumPy code and scale it
to multi-GPU systems with minimal code changes through Dask arrays containing
multiple CuPy arrays \cite{entschev2019}.

@@

% RG: this first sentence is duplicate with (b) three paragraphs up - on
%     purpose or not?
NumPy is a community of practice with a strong \emph{culture} of
employing software engineering practice to improve collaboration and
reduce error \cite{millman2014developing}.  This culture is not only
adopted by leaders in the project, but also enthusiastically taught to
newcomers. The NumPy team was early in adopting distributed revision
control and code review to improve collaboration on code, and
continuous testing that runs a large battery of automated tests for
every proposed change to NumPy.  The project has comprehensive,
high-quality documentation, integrated with the source
code \cite{vanderwalt2008scipy,harrington2008scipy,harrington2009scipy}.

@@

This culture of using best practices for producing reliable scientific software
has been adopted by the ecosystem of libraries that build on NumPy.
For example, in a recent award given by the Royal Astronomical Society to
Astropy, they state:
\begin{quotation}
\noindent\emph{The Astropy Project has provided hundreds of junior scientists
with experience in professional-standard software development practices
including use of version control, unit testing, code review and issue tracking
procedures. This is a vital skill set for modern researchers that is often
missing from formal university education in physics or astronomy.}
\end{quotation}
Community members explicitly work to address this lack of formal education
through courses and workshops
\cite{wilson-software-carpentry,hannay-scientific-software-survey,millman2018teaching}.
